---
title: "Survival Analysis P3"
categories: [Survival Analysis, Kaplan-Meier]
description: 'Self-Study Notes for Survival Analysis'
---

The goal of this post is to introduce the **Kaplan-Meier** estimate of the survival function. The Kaplan-Meier estimator makes no assumptions about the shape of the underlying survival distribution, making it a *non-parametric* and hence flexible to adapt to real world data.

If there were no censoring, this would be an easy task as we could directly use the ECDF $$\hat{F_n}(t) = \frac{1}{n} \sum_{i=1}^n I(t_i \le t) \implies \hat{S}_n(t) = 1 - \hat{F}_n(t)$$

# Derivation of the Kaplan Mier Estimate

To build up the Kaplan-Meier estimator, we assume that time is discrete.

Suppose we observe individual time points $t_0 = 0, t_1, t_2, \cdots$. These may be days, weeks, or any time unit relevant to the study. At each time $t_j$, an individual may experience the event of interest, while others continue on.

Next, define

-   $q_j$ be the probability of failing at time $t_j$, given that the subject survived up to that point, (ie. past time $t_{j-1}$
-   $p_j = 1 - q_j$ be the probability of surviving past time $t_j$, given survival up to time $t_{j-1}$

To express $p_j$ more formally, $$p_j = P(T > t_j| T > t_{j-1}) = \dfrac{P(T > t_j, T > t_{j-1})}{P(T > t_{j-1})} = \dfrac{P(T > t_j)}{P(T > t_{j-1})}$$ Recall that the survival function is defined as $$S(t) = P(T > t)$$

Hence, rewriting $p_j$ results in $$p_j = \dfrac{S(t_j)}{S(t_{j-1})} \tag{1}$$

We observe a recursive relationship $$S(t_0) = S(0) = 1$$ $$p_1 = \dfrac{S(t_1)}{S(t_0)} = S(t_1)$$ $$p_2 = \dfrac{S(t_2)}{S(t_1)}$$ $$p_2 = \dfrac{S(t_3)}{S(t_2)}$$ $$...$$ $$p_k = \dfrac{S(t_k)}{S(t_{k-1})}$$

Hence, $$p_1 \cdot p_2 \cdot p_3 \cdots p_k = S(t_1) \dfrac{S(t_2)}{S(t_1)} \dfrac{S(t_3)}{S(t_2)} \cdots  \dfrac{S(t_k)}{S(t_{k-1})} = S(t_k)$$

To estimate the survival function $S(t)$, we need to estimate each of the conditional survival probabilities $p_j$. At each time point $t_j$, define

-   $d_j$ as the number of events that occur exactly at time $t_j$
-   $n_j$ as the number of individuals in the risk set just before $t_j$

Note that anyone who was censored before time $t_j$ is no longer in the risk set.

We can then estimate the conditional probability of failure at time $t_j$ and the conditional probability of survival as $$\hat{q}_j = \dfrac{d_j}{n_j}$$ $$\hat{p}_j = 1 - \hat{q}_j = \dfrac{n_j - d_j}{n_j}$$

To estimate the overall survival probability at time $t_k$, we multiply all the conditional probabilities up to that time $$\hat{S}(t) = \prod_{t_j \le t} \left(1 - \frac{d_j}{n_j}\right)$$

# Greenwood's Formula

Test test

# Nelson-Aalen

# Application in R

## Fitting the model

::: callout-note
**Example: KM Estimate in R**.

Use R to compute the KM estimate of the survival function using the Veteran dataset.
:::

```{r}
#| echo: true
#| message: false
#| warning: false

library(survival)
library(ggplot2)

surv.veteran <- survfit(Surv(time, status) ~ trt, data = veteran) # Fitting model
# print(summary(surv.veteran), digits=4) # Summary model
```

## Plotting the survival function

```{r}
# Plotting KM survival curve by treatment group
km <- summary(surv.veteran)

df <- data.frame(
  time = km$time,
  surv = km$surv,
  group = factor(km$strata, 
                 levels = c("trt=1", "trt=2"),
                 labels = c("Standard", "Test"))
)

ggplot(df, aes(x = time, y = surv, color = group)) +
  geom_step() +
  labs(x = "Time (days)", y = "Survival Probability", color = "Group") +
  theme_minimal()
```
