---
title: "Survival Analysis P3: Kaplan Mier Estimate of the Survival Function"
categories: [Survival Analysis, Kaplan-Meier]
description: 'Self-Study Notes for Survival Analysis'
---

The goal of this post is to introduce the **Kaplan-Meier** estimate of the survival function. The Kaplan-Meier estimator makes no assumptions about the shape of the underlying survival distribution, making it a *non-parametric* and hence flexible to adapt to real world data.

Note that if there were no censoring, this would be an easy task as we could directly use the ECDF $$\hat{F_n}(t) = \frac{1}{n} \sum_{i=1}^n I(t_i \le t) \implies \hat{S}_n(t) = 1 - \hat{F}_n(t)$$

# Derivation of the Kaplan Mier Estimate

To build up the Kaplan-Meier estimator, we assume that time is discrete.

Suppose we observe individual time points $t_0 = 0, t_1, t_2, \cdots$. These may be days, weeks, or any time unit relevant to the study. At each time $t_j$, an individual may experience the event of interest, while others continue on.

Next, define

-   $q_j$ be the probability of failing at time $t_j$, given that the subject survived up to that point, (ie. past time $t_{j-1}$
-   $p_j = 1 - q_j$ be the probability of surviving past time $t_j$, given survival up to time $t_{j-1}$

To express $p_j$ more formally, $$p_j = P(T > t_j| T > t_{j-1}) = \dfrac{P(T > t_j, T > t_{j-1})}{P(T > t_{j-1})} = \dfrac{P(T > t_j)}{P(T > t_{j-1})}$$ Recall that the survival function is defined as $$S(t) = P(T > t)$$

Hence, rewriting $p_j$ results in $$p_j = \dfrac{S(t_j)}{S(t_{j-1})}$$

We observe a recursive relationship $$S(t_0) = S(0) = 1$$ $$p_1 = \dfrac{S(t_1)}{S(t_0)} = S(t_1)$$ $$p_2 = \dfrac{S(t_2)}{S(t_1)}$$ $$p_2 = \dfrac{S(t_3)}{S(t_2)}$$ $$...$$ $$p_k = \dfrac{S(t_k)}{S(t_{k-1})}$$

Hence, $$p_1 \cdot p_2 \cdot p_3 \cdots p_k = S(t_1) \dfrac{S(t_2)}{S(t_1)} \dfrac{S(t_3)}{S(t_2)} \cdots  \dfrac{S(t_k)}{S(t_{k-1})} = S(t_k)$$

To estimate the survival function $S(t)$, we need to estimate each of the conditional survival probabilities $p_j$. At each time point $t_j$, define

-   $d_j$ as the number of events that occur exactly at time $t_j$
-   $n_j$ as the number of individuals in the risk set just before $t_j$

Note that anyone who was censored before time $t_j$ is no longer in the risk set.

We can then estimate the conditional probability of failure at time $t_j$ and the conditional probability of survival as $$\hat{q}_j = \dfrac{d_j}{n_j}$$ $$\hat{p}_j = 1 - \hat{q}_j = \dfrac{n_j - d_j}{n_j}$$

Hence to estimate the overall survival probability at time $t_k$, we essentially multiply all the conditional survival probabilities at each observed failure time up to that time. So we obtain $$\hat{S}(t) = \prod_{t_j \le t} \left(1 - \frac{d_j}{n_j}\right) \tag{1}$$


# Greenwood's Formula

To quantify uncertainty around the survival estimate, we need its variance. 

We start from $$\hat{p}_j = 1 - \hat{q}_j = \dfrac{n_j - d_j}{n_j}$$ and recall this represents the probability of surviving past time $t_j$, given surviving past $t_{j-1}$. Here, $d_j =$ number of observed events at $t_j$ and $n_j =$ the number of people still in the risk set just before $t_j$.

In other words, $\hat{p}_j$ is just the proportion of people at risk who survived up to $t_j$. 

So if we define $X_j \sim$ Binomial$(n_j, p_j)$ to be the total number of survivors at $t_j$, our estimate becomes $\hat{p_j} = \frac{X_j}{n_j}$ and hence $$V(\hat{p_j}) = V(\frac{X_j}{n_j}) = \frac{p_j(1-p_j)}{n_j} \tag{1}$$

We wish to find the variance of $$\hat{S}(t) = \prod_{j \le t} \hat{p_j}$$ 

This alone is hard to work with directly. So we take logs:
$$\ln{\hat{S}(t)} = \sum_{j \le t}{\ln{\hat{p}_j}}$$
We proceed by applying the Delta Method
$$V\left( \ln \hat{p}_j \right)
= \left( \frac{d}{dp_j} \ln p_j \right)^2 \cdot V(\hat{p}_j)
= \left( \frac{1}{p_j} \right)^2 \cdot \frac{p_j (1 - p_j)}{n_j}
= \frac{1 - p_j}{p_j n_j}$$

Then we sum the variances on the log scale
$$V\left( \ln \hat{S}(t) \right)
= \sum_{j \le t} V\left( \ln \hat{p}_j \right)
= \sum_{j \le t} \frac{1 - p_j}{p_j n_j}$$

Simplifying using $1 - p_j = \frac{d_j}{n_j}$ and $\hat{p}_j = \frac{n_j - d_j}{n_j}$, we get
$$V\left[ \log \hat{S}(t) \right] = \sum_{t_j \leq t} \frac{d_j}{n_j (n_j - d_j)}$$
Lastly, we apply Delta Method once again to transform back to get Greenwood's Formula
$$V[\hat{S}(t)] \approx \hat{S}(t)^2 \cdot \sum_{t_j \leq t} \frac{d_j}{n_j(n_j - d_j)}$$

In R's `survival` package, the default confidence interval of the Kaplan Mier estimate uses this formula.

# Nelson-Aalen

The Nelson-Aalen estimate estimator of the cumulative hazard is:
$$\hat{H}_{\text{NA}}(t) = \sum_{t_j \leq t} \frac{d_j}{n_j}$$

Recall the Kaplan-Mier estimate is
$$\hat{S}_{\text{KM}}(t) = \prod_{t_j \leq t} \left(1 - \frac{d_j}{n_j}\right)$$


Using the approximation
$$\ln(1 - x) \approx -x$$ for small $x$, we have

$$\prod_{t_j \leq t} \left(1 - \frac{d_j}{n_j}\right) 
\approx \exp\left( -\sum_{t_j \leq t} \frac{d_j}{n_j} \right)$$

# Application in R
## Fitting the model

::: callout-note
**Example: KM Estimate in R**.

Use R to compute the KM estimate of the survival function using the Veteran dataset.
:::

```{r}
#| echo: true
#| message: false
#| warning: false

library(survival)
library(ggplot2)

surv.veteran <- survfit(Surv(time, status) ~ trt, data = veteran) # Fitting model
print(summary(surv.veteran), digits=4) # Summary model
```

For example, we can verify our KM estimate at $t_1 = 3$ is $1 - \frac{1}{69} = 0.98551$.


## Plotting the survival function

```{r}
# Plotting KM survival curve by treatment group
km <- summary(surv.veteran)

df <- data.frame(
  time = km$time,
  surv = km$surv,
  group = factor(km$strata, 
                 levels = c("trt=1", "trt=2"),
                 labels = c("Standard", "Test"))
)

ggplot(df, aes(x = time, y = surv, color = group)) +
  geom_step() +
  labs(x = "Time (days)", y = "Survival Probability", color = "Group") +
  theme_minimal()
```
